Using an SMT engine to generate Symbolic Automata

### 2019-03-07 2017-Xudong-SMT/ActaInformatica.tex
Acta Informatica style

\documentclass{eceasst}
\usepackage{etex}

@@

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
\usepackage{enumitem}
\setlist{nolistsep}
%\usepackage{indentfirst}
%%%\usepackage{amsmath,amssymb,amscd,mathrsfs}
% \usepackage{array,booktabs,arydshln,xcolor}
%
\usepackage{array,booktabs}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%\usepackage{epsfig,color,enumitem}
\usepackage{epsfig,enumitem}
\usepackage[%
    font={small,sf},
    labelfont=bf,
    format=hang,    
    format=plain,
    margin=0pt,
    %% width=0.8\textwidth,
    aboveskip=4pt,
    belowskip=-18pt,
]{caption}
\usepackage[list=true]{subcaption}

@@

% Volume frontmatter
% ==================
\input{eceasst/frontmatter}

\title{Using an SMT engine to generate Symbolic Automata}
\author{ Xudong Qin\autref{1}\sponsor{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai},
  Simon Bliudze\autref{2}, Eric Madelaine\autref{3}, Zechen Hou\autref{1}, Yuxin Deng\autref{1}, and Min Zhang\autref{4}\todoMargin{Should we remove Min ?}} 
\institute{\autlabel{1} \email{marsxd@gmail.com}\\
  \autlabel{4} \email{mzhang@sei.ecnu.edu.cn}\\
  Shanghai Key Laboratory of Trustworthy Computing, ECNU, China\par
  \autlabel{2} \email{simon.bliudze@inria.fr}\\
  INRIA Lille -- Nord Europe, 40 avenue Halley, 59650 Villeneuve d'Ascq,
France\par
  \autlabel{3} \email{eric.madelaine@inria.fr}\\
  Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia Antipolis, France
}
\date{}                                           % Activate to display a given date or no date

@@

@@

\abstract{Open pNets are used to model the behaviour of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called ``architectures''
for BIP systems, that have been used in the framework of an ESA project and to specify the control software of
a nanosatellite at the EPFL Space Engineering Center. We use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties; we give 2 examples, a safety and a liveness
property.
\todoMargin{Add new contrib}}
\keywords{Networks of Synchronised Automata, Symbolic Behavioral Semantics, Compositional Analysis of Software Components, SMT}

\maketitle

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We define the open automaton generation algorithm. We
  implemented a full working prototype, within the 
  VerCors platform. In the process, we improved the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal with the full 
  model, including management of variables and assignments.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
    \item \ERIC{we define a strategy checking satisfiability of open transitions in the course the residual algorithm, and show that this ensures that we still generate the reachable part of the open automaton, and that what is generated is minimal with respect to the decidability of predicate decidability} 
\item We show the interest of this approach \ERIC{on various use-cases from different settings}, including an
    industry-inspired case-study, namely one architectural pattern
    extracted (and extended) from the BIP specification of a nanosatellite on-board
    software.
\end{itemize}

@@

\begin{proof}
  \noteEM{I tentatively copied the proof here, from RR-8898. We should
    check if it is sufficient with this version of the algorithm, or
    if it can be make more precise (some added analysis of the
    behavior of matching/combining/filetring ?)}
  The possible set of states of the open-automaton is the cartesian
  product of the states of its leaves $pLTS_i^{i\in L}$, that is finite by
  hypothesis. So the top-level residual loop of Algorithm 1 terminates provided each 
  iteration terminates. The enumeration of open-transitions in step 2b is bounded by the
  number of applications of rules Tr2 on the structure of the
  pNet tree, with a finite number of synchronisation vectors applying
  at each node the number of global open transition is finite. Similarily rule Tr1 is 
  applied finitely if the number
  of transitions of each pLTS is finite. So we get finitely many deduction trees, and
  open-transitions which ensures that each internal loop of Algorithm 1 terminates.
  \end{proof}

@@

\subsection{Result for the running example}
\label{section:resultOA}
\noteEM{This is the original example from the Avocs paper}
For this example, the tool builds 184 open transitions, whereof 173
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 11 open
transitions, is shown in Fig.~\ref{schema:resultOA}.

@@

\subsection{Extended version of the FailureTimer BIP Architecture}
\label{section:resultTimerInterval}
\noteEM{Here we have extended the FailureTimer architecture using time intervals,
  that makes more interesting the timeout/reset temporal behavior}

>>>>

\documentclass[actinf,draft]{svjour}
%\documentclass[draft]{svjour3}
\usepackage{etex}

@@

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
\usepackage{enumitem}
\setlist{nolistsep}
%\usepackage{indentfirst}
%%%\usepackage{amsmath,amssymb,amscd,mathrsfs}
% \usepackage{array,booktabs,arydshln,xcolor}
%
\usepackage{array,booktabs}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%\usepackage{epsfig,color,enumitem}
\usepackage{epsfig,enumitem}
%% \usepackage[%
%%     font={small,sf},
%%     labelfont=bf,
%%     format=hang,    
%%     format=plain,
%%     margin=0pt,
%%     %% width=0.8\textwidth,
%%     aboveskip=4pt,
%%     belowskip=-18pt,
%% ]{caption}
%% \usepackage[list=true]{subcaption}

@@

% Volume frontmatter
% ==================
% \input{eceasst/frontmatter}

\title{Using an SMT engine to generate Symbolic Automata}
\author{ Xudong Qin\inst{1},
  Simon Bliudze\inst{2} \and Eric Madelaine\inst{3} \and Zechen
  Hou\inst{1} \and Yuxin Deng\inst{1} \and Min Zhang\inst{4}}%\todoMargin{Should we remove Min ?}} 
%% \institute{\autlabel{1} \email{marsxd@gmail.com}\\
%%   \autlabel{4} \email{mzhang@sei.ecnu.edu.cn}\\
%%   Shanghai Key Laboratory of Trustworthy Computing, ECNU, China\par
%%   \autlabel{2} \email{simon.bliudze@inria.fr}\\
%%   INRIA Lille -- Nord Europe, 40 avenue Halley, 59650 Villeneuve d'Ascq,
%% France\par
%%   \autlabel{3} \email{eric.madelaine@inria.fr}\\
%%   Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia Antipolis, France
%% }
\institute{Shanghai Key Laboratory of Trustworthy Computing, ECNU,
  China, \email{marsxd@gmail.com, zechen@sei.ecnu.edu.cn,
  yxdeng@sei.ecnu.edu.cn, mzhang@sei.ecnu.edu.cn}
  \and INRIA Lille -- Nord Europe, 40 avenue Halley, 59650 Villeneuve
  d'Ascq, \email{simon.bliudze@inria.fr}
  \and Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia
  Antipolis, France, \email{eric.madelaine@inria.fr}}

% \date{}                                           % Activate to display a given date or no date

@@

\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}
\maketitle

@@

\abstract{Open pNets are used to model the behaviour of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called ``architectures''
for BIP systems, that have been used in the framework of an ESA project and to specify the control software of
a nanosatellite at the EPFL Space Engineering Center. We use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties; we give 2 examples, a safety and a liveness
property.
\TODO{Add new contrib}
\keywords{Networks of Synchronised Automata, Symbolic Behavioral Semantics, Compositional Analysis of Software Components, SMT}
}

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We define the open automaton generation algorithm. We
  implemented a full working prototype, within the 
  VerCors platform. In the process, we improved the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal with the full 
  model, including management of variables and assignments.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
    \item \ERIC{we define a strategy checking satisfiability of open transitions in the course the residual algorithm, and show that this ensures that we still generate the reachable part of the open automaton, and that what is generated is minimal with respect to the decidability of predicate satisfiability} 
\item We show the interest of this approach \ERIC{on various use-cases from different settings}, including an
    industry-inspired case-study, namely one architectural pattern
    extracted (and extended) from the BIP specification of a nanosatellite on-board
    software.
\end{itemize}

@@

\begin{proof}
  \TODO{I tentatively copied the proof here, from RR-8898. We should
    check if it is sufficient with this version of the algorithm, or
    if it can be make more precise (some added analysis of the
    behavior of matching/combining/filetring ?)}
  The possible set of states of the open-automaton is the cartesian
  product of the states of its leaves $pLTS_i^{i\in L}$, that is finite by
  hypothesis. So the top-level residual loop of Algorithm 1 terminates provided each 
  iteration terminates. The enumeration of open-transitions in step 2b is bounded by the
  number of applications of rules Tr2 on the structure of the
  pNet tree, with a finite number of synchronisation vectors applying
  at each node the number of global open transition is finite. Similarily rule Tr1 is 
  applied finitely if the number
  of transitions of each pLTS is finite. So we get finitely many deduction trees, and
  open-transitions which ensures that each internal loop of Algorithm 1 terminates.
  \end{proof}

@@

\subsection{Result for the running example}
\label{section:resultOA}
\TODO{This is the original example from the Avocs paper}
For this example, the tool builds 184 open transitions, whereof 173
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 11 open
transitions, is shown in Fig.~\ref{schema:resultOA}.

@@

\subsection{Extended version of the FailureTimer BIP Architecture}
\label{section:resultTimerInterval}
\TODO{Here we have extended the FailureTimer architecture using time intervals,
  that makes more interesting the timeout/reset temporal behavior}
